Bandwidth-Aware Reconfigurable Memory Design with Hybrid Memory Technologies

### 2011-04-07 paper.tex

\begin{abstract}
  In future chip-multiprocessor~(CMP) design, memory bandwidth is a potential
  bottleneck to system performance. Emerging non-volatile memory technologies,
  such as spin-torque-transfer memory~(STT-RAM), resistive memory~(RRAM), and
  embedded DRAM~(eDRAM), are promising solutions as on-chip memories for CMPs.
  In this paper, we propose a bandwidth-aware reconfigurable on-chip
  memory~(BAROM) design with hybrid memory technologies. We demonstrate a method
  to dynamically reconfigure the number of shared memory levels and capacity of
  each level based on statistical prediction. With a set of both multithreaded
  and multiprogrammed applications, we evaluate system performance obtained with
  our method. The experimental results show that the proposed reconfigurable
  hybrid memory design improves the overall system throughput by \% compared to
  pure SRAM memory design. In addition, our design leads to \% throughput
  increase and \% dynamic power reduction compared to hybrid but fixed memory
  hierarchy design.

@@

\begin{itemize}
\item Bandwidth-aware hybrid on-chip memory hierarchy design.
\item Hardware support that enables prediction of cache performance on the
  different sized caches.
\item OS scheduler support to make use of the prediction capability and
  appropriately schedule applications on to core with suitable cache capacity.

@@

Based on our memory technology exploration, we now provide an overview of BAROM
design. While many design methods envolved with new memory technologies endeavor
to reduce the off-chip memory access latency, this work focuses on minimize
off-chip bandwidth requirement by employing hybrid on-chip memory and
reconfiguration. For each level of memory, we select the memory technology that
has the highest provided bandwidth within given capacity. On top of such
hardware hierarchy design, we apply dynamic memory system reconfiguration with
the bandwidth demand of a specific set of applications as the major metric. One
desirable characteristic of BAROM is that it provides prioritized access to the
residual memory capacity available at runtime instead of sharing it among all
data flows. By having control over the residual capacity, we can better utilize
the limited resources to enhance system performance. The hardware
reconfiguration is further facilitated by software prediction engine. Rather
than conventional last value predictors and table-based predictors, we employ a
probablity-based predictor that can achieve higher accuracy with small
performance overhead. The rest of this section describes the BAROM design, which
is composed of on-chip memory hierarchy hardware design, the reconfiguration
mechanism, and the predictor engine.

\subsection{Overview}

@@

The major system parameter to be reconfigured is the capacity of each memory
level. At each time interval, we determine the upper bound of the capacity of
memory level-i ($s_{iu}$) by mapping the demand bandwidth (DBW) of a specific
application to the system's provided bandwidth curve, i.e.,
$s_{iu}=f^{-1}(\textrm{DBW})$ where $f(x)$ represents the provided bandwidth
curve of the on-chip memory system stored in the look-up-table in a reserved
non-volatile memory space. The demand bandwidth for each memory level is
generated using the prediction engine presented in section~\ref{sec:predict}. In
previous sections, we show that the memory with smaller capacity provides higher
bandwidth. Theoritically, the higher bandwidth provided by the memory system
leads to higher performance (in terms of throughput) and power consumption of
throughput computing applications~\cite{throughput-vs-bandwidth}. Consequently,
we define a lower bound of memory capacity as
$s_{il}=f^{-1}(\textrm{DBW}*(1+\sigma))$, where $\sigma$ is a pre-defined
threshold to make sure possible provided bandwidth increase without much power
overhead. The capacity of memory level-i ($s_i$) is therefore selected to
satisfy $s_{il}\le s_i\le s_{iu}$. Furthermore, the rest of the capacity can be
reconfigured to become level-i.5 if necessary. We describe the reconfiguration
algorithm as follows.

@@

statistical prediction

@@

>>>>

\begin{abstract}
  In future chip-multiprocessor~(CMP) design, memory bandwidth is a potential
  bottleneck to system performance. Emerging non-volatile memory technologies,
  such as spin-torque-transfer memory~(STT-RAM), resistive memory~(RRAM), and
  embedded DRAM~(eDRAM), are promising solutions as on-chip memories for CMPs.
  In this paper, we propose a bandwidth-aware reconfigurable on-chip
  memory~(BAROM) design with hybrid memory technologies. We demonstrate a method
  to dynamically reconfigure the number of shared memory levels and capacity of
  each level based on statistical prediction. With a set of both multithreaded
  and multiprogrammed applications, we evaluate system performance obtained with
  our method. The experimental results show that the proposed reconfigurable
  hybrid memory design improves the overall system throughput by \% compared to
  pure SRAM memory design. In addition, our design leads to \% throughput
  increase compared to hybrid but fixed memory hierarchy design.

@@

\begin{itemize}
\item Bandwidth-aware hybrid on-chip memory hierarchy design...
\item Reconfiguration...
\item Prediction engine...

@@

Based on our memory technology exploration, we now provide an overview of BAROM
design. While many design methods envolved with new memory technologies endeavor
to reduce the off-chip memory access latency, this work focuses on minimize
off-chip bandwidth requirement by employing hybrid on-chip memory and
reconfiguration. For each level of memory, we select the memory technology that
has the highest provided bandwidth within given capacity. The total capacity of
each level is then partitioned to subsets of ways with different provided
bandwidths. On top of such hardware hierarchy design, we apply dynamic memory
system reconfiguration with the bandwidth demand of a specific set of
applications as the major metric. One desirable characteristic of BAROM is that
it provides prioritized access to the residual memory capacity available at
runtime instead of sharing it among all memory accesses. By having control over
the residual capacity, we can better utilize the limited resources to enhance
system performance. The hardware reconfiguration is further facilitated by a
prediction engine. Rather than conventional last value or history table based
predictors, we employ a probablity-based statistical predictor which can achieve
higher accuracy with small performance overhead. The rest of this section
describes the BAROM design, which is composed of on-chip memory hierarchy
design, the reconfiguration mechanism, and the prediction engine.

\subsection{Memory Hierarchy Design}

@@

The major system parameter to be reconfigured is the capacity of each memory
level. At each time interval, we determine the upper bound of the capacity of
memory level-i ($s_i^u$) by mapping the demand bandwidth (DBW) of a specific
application to the system's provided bandwidth curve, i.e.,
$s_i^u=f^{-1}(\textrm{DBW})$ where $f(x)$ represents the provided bandwidth
curve of the on-chip memory system stored in the look-up-table in a reserved
non-volatile memory space. The demand bandwidth for each memory level is
generated using the prediction engine presented in section~\ref{sec:predict}. In
previous sections, we show that the memory with smaller capacity provides higher
bandwidth. Theoritically, the higher bandwidth provided by the memory system
leads to higher performance (in terms of throughput) and power consumption of
throughput computing applications~\cite{throughput-vs-bandwidth}. Consequently,
we define a lower bound of memory capacity as
$s_i^l=f^{-1}(\textrm{DBW}*(1+\sigma))$, where $\sigma$ is a pre-defined
threshold to make sure possible provided bandwidth increase without much power
overhead. The capacity of memory level-i ($s_i$) is therefore selected to
satisfy $s_i^l\le s_i\le s_i^u$. Furthermore, the rest of the capacity can be
reconfigured to become level-i.5 if necessary. We describe the reconfiguration
algorithm as follows.

<pseudocode>
if (si < stotali/2): 
   if (si+1 <= stotali-si):
      si.5 = stotali-si;

else:
   si+1 = ...;

@@

Memory reconfiguration relies on accurate predictions of demand bandwidth of a
workload in response to the dynamically varying application characteristics.
Conventional last value or table based predictors can model neither long range
patterns of an application nor patterns with variable lengths. In this work, we
employ a statistical predictor to support for the BAROM design. The basic idea
of the predictor is borrowed from the n-Gram models, which are typically used by
speech and natural language processing~\cite{smoothing-tech}. The n-Gram models
are usually formulated as a probability distribution $p(s)$ over a strings $s$,
and attemp to reflect how frequently the string occurs as a sentense. Prediction
on a reasonable combination of strings is then generated based on the
probability distribution. Our predictor employs the same basic idea, but
simplifies the model for performance considerations. In our model, each demand
bandwidth sample abtained in a time interval is a string, and a given length
(the ``order'') of samples is stored in a table as a pattern. We calculate the
probability of each pattern at each time interval, and predict the demand
bandwidth for the next time interval based on the probabilities.

Figure~\ref{fig:predictor} shows the hardware components of our prediction
engine, which includes a pattern look-up-table, a probability vector, and a set
of counters.

The algorithm of the predictor is described in Figure~\ref{algo:predictor}.

hardware: global predictor sitting before caches
statistical prediction

@@

1. nvsim memory bandwidth vs. way-configuration -> reasonable fast ways and slow
ways for reconfiguration

2. predictor accuracy vs. benchmarks

3. performance of instruction throughput vs. benchmarks

------------------------------------
